A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-03-05 analysis/paper/paper.Rmd
update to figures

### Method selection
We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018]. We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues). Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).

>>>>

### Method selection
We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`, als), by searching in literature for "trajectory inference" and "pseudotemporal ordering", and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018]. We welcome any contributions by creating an issue at [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues). Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).

### 2018-03-05 analysis/paper/paper_latex.Rmd
update to figures

# Introduction
<!-- §1: Single-cell omics and TI --->
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. When cells are profiled from a population in which cells are at different unknown points in the dynamic process, this process  processes can be modelled using trajectory inference (TI) methods  [@cannoodt_computational_2016]. These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex. TI methods offer an unbiased and transcriptome-wide understanding of dynamic process [@tanay_scaling_2017], and thus allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a bifurcation [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

```{r}
add_sfig(
figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
"methods_over_time", 
"New TI methods over time. **a** Number of methods published or in preprint. **b** Number of methods which can handle a particular type of trajectory."
)
```

@@

```{r, results='asis' }
add_table(
"implementations",
read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
"Overview of current trajectory inference methods, and whether they were included in this current evaluation study"
)
```
<!-- §3: Topology inference -->
One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. The focus of these methods was therefore mainly to correctly order the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows a method to be more broadly applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`). Most noticeable of these methods is AGA[@wolf_graphabstractionreconciles_2017], which is the only TI method currently able to predict disconnected graphs.

@@

```{r}
add_sfig(
figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
"implementation_components", 
"TI methods frequently share steps"
)
```

@@

```{r}
add_sfig(
        figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
"dataset_characterisation",  
"Characteristics of the real datasets used for the evaluation"
)
```

```{r}
add_sfig(
        figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
"synthetic_data_generation",  "Synthetic data generation workflow"
)
```

We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation. We scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review. This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets.

Using both the benchmark and quality control results, we will finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.

@@

```{r}
add_sfig(
        {figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview')}, 
"extended_evaluation_overview", "Extended overview ", 15, 7.5
)
```

@@

<!-- §3: real versus synthetic -->
A high correlation can be observed between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`). This confirms both the relevance of the synthetic data and the accuracy of the real datasets.

```{r}
add_sfig(
figure_file('real_synthetic_comparison.tmp.pdf', "5-optimise_parameters/5-real_synthetic_comparison"), 
"real_synthetic_comparison", 
"Comparison between performance on real and synthetic datasets across trajectory types."
)
```

<!-- §4: Compare metrics -->
As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is unsurprising to observe differences in rankings of TI methods amongst the different metrics. While there is no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also seed to obtain a higher RF MSE score (`r ref("sfig", "metric_comparison")`). 
In addition, methods able to detect more complex trajectory types seem to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology. This issue will also be investigated in more detail in Section SEC_TRAJTYPE.

@@

```{r}
add_sfig({
        figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology")
}, 
"trajtype_handle_comparison", 
""
)
```

```{r}
add_sfig({
        figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology")
}, 
"linear_split_comparison", 
"Linear methods have a higher correlation score on non-linear datasets. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods. "
)
```

@@

`r anchor("fig", "topology_complexity_comparison")`
![](`r figure_file("topology_complexity_comparison.tmp.pdf","8-compare_topology")`)
`r ref("fig", "topology_complexity_comparison")`

```{r}
add_sfig({
        figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology")
}, 
"topology_sensitivity", 
""
)
```

@@

```{r}
add_sfig({
        figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information")
}, 
"required_priors_performance_comparison", 
"Comparison of method performance "
)
```

@@

```{r}
add_stable(
read_rds(figure_file("component_importance.rds","6-developer_guidelines"))[[params$table_format]], 
"component_importance", 
"Importance values for algorithmic components"
)
```

@@

```{r}
add_stable(
read_rds(figure_file("qc_checks.rds","4-method_characterisation"))[[params$table_format]], 
"qc_checks", 
"Scoring scheme for method quality control"
)
```

@@

`r anchor("fig", "implementation_qc_overview")`
![](`r figure_file("implementation_qc_overview.tmp.pdf","4-method_characterisation")`)
`r ref("fig", "implementation_qc_overview")` Overview of the quality control results for every method.

```{r}
add_sfig(
figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
"qc_over_time", 
"Quality control scores and number of citations over time"
)
```
We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`). We could also not find any quality aspect which was significantly associated with method performance.

```{r}
add_sfig(
figure_file("qc_performance_comparison.tmp.pdf","5-optimise_parameters/6-qc_performance_comparison"), 
"qc_performance_comparison", 
"Comparison between method qc (x-axis) and method performance (y-axis)."
)
```

@@

### Method selection
We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018]. We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynalysis/issues). Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).

### Method input

As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method. Furthermore, when required, we also provided a maximum of 7 types of prior information. This prior information was extracted from the gold/silver standards as follows:

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges. For ties, we chose one random cell. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.

@@

```{r}
add_sfig(
        {figure_file('wrapping.tmp.pdf', '4-method_characterisation')},
"wrapping",  "Our wrapping workflow", 15, 7.5
)
```

@@

# create functions to list the methods within a particular conversion category
list_sentence <- function(x) {
  pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))
}
conversion <- function(conversion_oi) {
 methods_evaluated %>% 
   filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% 
   mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>%
   pull(text) %>% 
   list_sentence()
}

```

@@

```{r}
add_stable(
read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
"datasets", 
"Real datasets used in this study"
)
```

@@

```{r}
add_sfig({
read_rds(figure_file("modulenets.rds", experiment_id="2-dataset_characterisation/3-synthetic"))
}, 
"modulenets", 
"Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
15, 
5
)
```

@@

```{r}
add_stable(
read_rds(figure_file("samplers.rds","2-dataset_characterisation/3-synthetic"))[[params$table_format]], 
"samplers", 
"Distributions from which each parameter in the thermodynamic model was sampled"
)
```

@@

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernouilli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.

@@

Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory toplogy is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable. To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

@@

```{r}
add_sfig({
read_rds(figure_file("score_aspects.rds", experiment_id="3-score_rules/5-score_aspects"))
}, 
"score_aspects", 
"Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.", 
15, 
8
)
```

@@

As a third independent score, we assessed the similarity between the milestone network topology. We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that A->A would be converted A->B->C->D, by adding an intermediate node to linear networks. Because we are interested , the network was made undirected. Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no scalable solutions [@bahiense_maximumcommonedge_2012]. We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes. To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

```{r}
add_sfig({
figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip")
}, 
"edge_flip_comparison", 
paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
15, 
5
)
```

### Aggregation of scores
```{r}
add_sfig({
figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview")
}, 
"aggregation_of_results", 
"Aggregation methodology of metric scores", 
15, 
5
)
```

@@

### Pipeline
* Look at dyneval methods to write this paragraph

@@

```{r}
qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation"))$references %>% 
  str_split("[, \n]+") %>% 
  unlist() %>% 
  unique() %>% 
  discard(is.na) %>%
  glue::collapse(., ";") %>%
  paste0("[", ., "]")
```

@@

### Trajectory types and topology complexity
It's already a bit better, though could be improved still!
Expand, but maybe it should be moved to discussion anyway
😉
to do: add metric figure
needs to be updated with waypoints
this is not used anymore
TODO
More to add here?
* Look at dynmethods::execute_method to write this paragraph
Are there more exectuion details?

>>>>

# Introduction
<!-- §1: Single-cell omics and TI --->
Single-cell -omics technologies now make it possible to model biological systems more accurately than ever before. One area where single-cell data has been particularly useful is in the study of cellular dynamic processes, such as the cell cycle, cell differentiation and cell activation [@etzrodt_quantitativesinglecellapproaches_2014]. Such dynamic processes can be computationally modelled using trajectory inference (TI) methods (also known as pseudemporal ordering methods), by profiling cells from a population in which the cells are at different unknown points in the dynamic process  [@cannoodt_computational_2016]. These methods computationally order the cells along a trajectory topology, which can be linear, bifurcating, or more complex tree or graph structures. Because TI methods offer an unbiased and transcriptome-wide understanding of a dynamic process [@tanay_scaling_2017],  they allow the objective identification of new (primed) subsets of cells [@schlitzer_identificationcdc1cdc2committed_2015], delineation of a differentiation tree [@velten_humanhaematopoieticstem_2017; @see_mappinghumandc_2017] and inference of regulatory interaction responsible for a one or more bifurcations [@aibar_scenicsinglecellregulatory_2017]. 
Current applications of TI focus on specific subsets of cells, but ongoing efforts to construct transcriptomic catalogues of whole organisms [@regev_scienceforumhuman_2017; @han_mappingmousecell_2018] underline the urgency for accurate, scalable[@aibar_scenicsinglecellregulatory_2017; @angerer_singlecellsmake_2017] and user-friendly TI methods.

```{r}
add_sfig(
  figure_file("methods_timeline.tmp.pdf", "4-method_characterisation"),
  "methods_over_time", 
  "New TI methods over time. a) Number of methods published or in preprint. b) Number of methods which can handle a particular type of trajectory."
)
```

@@

```{r, results='asis' }
add_table(
  "implementations",
  read_rds(figure_file("implementations_table.rds", "4-method_characterisation")),
  "Overview of current trajectory inference methods, and whether they were included in this current evaluation study."
)
```
<!-- §3: Topology inference -->
One distinctive characteristic of TI methods is whether the topology of the trajectory is inferred computationally, or was fixed by design.
Early TI methods typically fixed the topology algorithmically (e.g. linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]) or through parameters provided by the user [@trapnell_dynamicsregulatorscell_2014; @matsumoto_scoupprobabilisticmodel_2016]. The focus of these methods was therefore mainly to correctly order the cells along this fixed topology. 
Other methods attempt to infer the topology computationally, which increases the difficulty of the problem at hand, but allows a method to be more broadly applicable on more use cases. Methods that perform topology inference are still in the minority, though current trends suggest this will soon change (`r ref("sfig", "methods_over_time", "c")`). Most noticeable of these methods is AGA[@wolf_graphabstractionreconciles_2017], which is the only TI method currently able to predict disconnected graphs.

@@

```{r}
add_sfig(
  figure_file("implementation_components.tmp.pdf","4-method_characterisation"), 
  "implementation_components", 
  "TI methods frequently share steps."
)
```

@@

```{r}
add_sfig(
  figure_file('dataset_characterisation.tmp.pdf', '2-dataset_characterisation'),
  "dataset_characterisation",  
  "Characteristics of the real datasets used for the evaluation."
)
```

```{r}
add_sfig(
  figure_file('synthetic_data_generation.tmp.pdf', "2-dataset_characterisation/3-synthetic"),
  "synthetic_data_generation",
  "Synthetic data generation workflow."
)
```

We not only assessed the accuracy of the predictions made by a trajectory, but also investigated the quality of the method's implementation. To do this, we scored each method using a checklist of important scientific and software development practices, including software packaging, documentation, automated code testing, and peer review. This allowed us to rank each method based on its user friendliness, developer friendliness, and potential broad applicability on new unseen datasets. Finally, using both the benchmark results and quality control, we finally present a flow chart with practical guidelines for selecting the most appropriate TI method for a given use case.

@@

```{r}
add_sfig(
  figure_file('expanded_ti_evaluation_overview_v1.tmp.pdf', 'evaluation_overview'),
  "extended_evaluation_overview",
  "Extended overview", 
  15, 
  7.5
)
```

@@

<!-- §3: real versus synthetic -->
We observed a high correlation (0.7-0.9) between results originating from real datasets versus those originating from synthetic datasets (`r ref("sfig", "real_synthetic_comparison")`). This confirms both the relevance of the synthetic data and the accuracy of the gold standard in the real datasets. However, this correlation was lower for converging and multifurcating datasets, potentially because only a small number of real datasets were available which contain such topologies (`r ref("stable", "datasets")`).

```{r}
add_sfig(
  figure_file(
    "real_synthetic_comparison.tmp.pdf",
    "5-optimise_parameters/5-real_synthetic_comparison"
  ), 
  "real_synthetic_comparison", 
  "Comparison between performance on real and synthetic datasets across trajectory types."
)
```

<!-- §4: Compare metrics -->
As the different metrics were selected to assess the correctness of a trajectory in various approaches, it is expected to observe differences in rankings of TI methods amongst the different metrics. While we saw no direct link between the edge flip scores versus the correlation or RF MSE metric, methods that obtain a high correlation score also tended to obtain higher RF MSE scores (`r ref("sfig", "metric_comparison")`). In addition, methods able to detect more complex trajectory types seemed to obtain higher edge flip scores, in comparison to methods whose trajectory topology was fixed to a simple trajectory topology. We will explore this issue in more detail in a further section SEC_TRAJTYPE.

@@

```{r}
add_sfig(
  figure_file('trajtype_handle_comparison.tmp.pdf', "8-compare_topology"), 
  "trajtype_handle_comparison", 
  "ADD CAPTION"
)
```

```{r}
add_sfig(
  figure_file('linear_split_comparison.tmp.pdf', "8-compare_topology"), 
  "linear_split_comparison", 
  "Linear methods have a higher correlation score on non-linear datasets. Methods were split in linear methods (those which can either only detect cyclic or linear trajectories) and non-linear methods."
)
```

@@

`r anchor("fig", "topology_complexity_comparison")`
![](`r figure_file("topology_complexity_comparison.tmp.pdf","8-compare_topology")`)

`r ref("fig", "topology_complexity_comparison")` **INSERT CAPTION.** Caption subdetails.

```{r}
add_sfig(
  figure_file('topology_sensitivity.tmp.pdf', "8-compare_topology"), 
  "topology_sensitivity", 
  "ADD CAPTION"
)
```

@@

```{r}
add_sfig(
  figure_file('required_priors_performance_comparison.tmp.pdf', "10-prior_information"), 
  "required_priors_performance_comparison", 
  "Comparison of method performance."
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "component_importance.rds",
    "6-developer_guidelines"
  ))[[params$table_format]], 
  "component_importance", 
  "Importance values for algorithmic components."
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "qc_checks.rds",
    "4-method_characterisation"
  ))[[params$table_format]], 
  "qc_checks", 
  "Scoring scheme for method quality control."
)
```

@@

`r anchor("fig", "implementation_qc_overview")`
![](`r figure_file("implementation_qc_overview.tmp.pdf","4-method_characterisation")`)

`r ref("fig", "implementation_qc_overview")` **Overview of the quality control results for every method.**

```{r}
add_sfig(
  figure_file("qc_over_time.tmp.pdf","4-method_characterisation"), 
  "qc_over_time", 
  "Quality control scores and number of citations over time."
)
```
We observed no clear relation between method quality and method performance (`r ref("fig", "results_overview")` and `r ref("sfig", "qc_performance_comparison")`). We could also not find any quality aspect which was significantly associated with method performance.

```{r}
add_sfig(
  figure_file(
    "qc_performance_comparison.tmp.pdf",
    "5-optimise_parameters/6-qc_performance_comparison"
  ), 
  "qc_performance_comparison", 
  "Comparison between method qc (x-axis) and method performance (y-axis)."
)
```

@@

### Method selection
We gathered a list of `r sum(methods$is_ti, na.rm=TRUE)` trajectory inference methods (`r ref("table", "implementations")`), by searching in literature for "trajectory inference" and "pseudotemporal ordering" and based on two existing lists found online [@davis_awesomesinglecelllistsoftware_2018; @gitter_singlecellpseudotimeoverviewalgorithms_2018]. We believe this to be the most complete list to data, but are welcome to include additional methods through e-mail or an issue on our github [https://github.com/dynverse/dynalysis/issues](github.com/dynverse/dynverse/issues). Methods were excluded from the evaluation based on several criteria: `r glue::collapse(glue::glue("({non_inclusion_reasons$footnote}) {non_inclusion_reasons$long}"), ", ")`. This resulted in the inclusion of `r sum(methods$evaluated)` methods in the evaluation (`r ref("table", "implementations")`).

### Method input
As input, we provided for each method either the raw count data (after cell and gene filtering) or normalised expression values, based on the description in the methods documentation or from the study describing the method. Furthermore, when required, we also provided a maximum of 7 types of prior information. This prior information was extracted from the gold/silver standards as follows:

* **Start cells** The identity of one or more start cells. For both real and synthetic data, a cell was chosen which was the closest (in geodesic distance) from each milestone with only outgoing edges. For ties, one random cell was chosen. For cyclic datasets, a random cell was chosen.
* **End cells** The identity of one or more end cells. Similar as the start cells, but now for every state with only ingoing edges.
* **# end states** Number of terminal states. Number of milestones with only ingoing edges.
* **Grouping** For each cell a label to which state/cluster/branch it belongs. For real data, the states from the gold/silver standard. For synthetic data, each milestone was seen as one group, and cells were assigned to its closed milestone.
* **# branches** Number of branches/intermediate states. For real data, the number of states in the gold/silver standard. For synthetic data, the number of milestones.
* **Time course** For each cell a time point from which it was sampled. If available, directly extracted from the gold standard. For synthetic data: the simulation time.

@@

```{r}
add_sfig(
  figure_file('wrapping.tmp.pdf', '4-method_characterisation'),
  "wrapping", 
  "Our wrapping workflow.", 
  15, 
  7.5
)
```

@@

# create functions to list the methods within a particular conversion category
list_sentence <- function(x) {
  pritt(glue::collapse(x[1:length(x) - 1],", "), " and ", last(x))
}
conversion <- function(conversion_oi) {
 methods_evaluated %>% 
   filter(map_lgl(conversion_split, ~ conversion_oi %in% .)) %>% 
   mutate(text = paste0(method_name, ifelse(conversion_special, "\\*", ""))) %>%
   pull(text) %>% 
   list_sentence()
}
```

@@

```{r}
add_stable(
  read_rds(figure_file("datasets.rds","2-dataset_characterisation"))[[params$table_format]], 
  "datasets", 
  "Real datasets used in this study."
)
```

@@

```{r}
add_sfig(
  read_rds(figure_file(
    "modulenets.rds",
     experiment_id = "2-dataset_characterisation/3-synthetic"
   )), 
  "modulenets", 
  "Module networks, milestone networks and the module dynamics for each of the different types of synthetic data", 
  15, 
  5
)
```

@@

```{r}
add_stable(
  read_rds(figure_file(
    "samplers.rds",
    "2-dataset_characterisation/3-synthetic")
  )[[params$table_format]], 
  "samplers", 
  "Distributions from which each parameter in the thermodynamic model was sampled."
)
```

@@

For each dataset we sampled the same number of cells as were present in the reference real dataset, limited to the simulation steps after burn-in. Next, we used the Splatter package [@zappia_splattersimulationsinglecell_2017] to estimate the different characteristics of a real dataset, such as the distributions of average gene expression, library sizes and dropout probabilities. We used Splatter to simulate the expression levels $\lambda_{i,j}$ of housekeeping genes $i$ (to match the number of genes in the reference dataset) in every cell $j$. These were combined with the expression levels of the genes simulated within a trajectory. Next, true counts were simulated using $Y'_{i,j} \sim \text{Poi}(\lambda_{i,j})$. Finally, we simulated dropouts by setting true counts to zero by sampling from a Bernoulli distribution using a dropout probability $\pi^D_{i,j} =\frac{1}{1+e^{-k(\text{ln}(\lambda_{i,j})-x_0)}}$.

@@

Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory topology is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable. To estimate a cell's position during a simulation within the trajectory topology, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory topologies, the number of possible milestones a backbone could progress through was limited to 20.

@@

```{r}
add_sfig({
  read_rds(figure_file(
    "score_aspects.rds", 
    experiment_id="3-score_rules/5-score_aspects"))
  }, 
  "score_aspects", 
  "Different scores assess different aspects of the correspondence between a prediction and gold standard trajectory.",
  15, 
  8
)
```

@@

As a third independent score, we assessed the similarity between the milestone network topology. We first simplified each network, by merging consecutive linear edges into one edge, by adding new milestones within self loops such that $A \rightarrow A$ would be converted $A \rightarrow B \rightarrow C \rightarrow D$, by adding an intermediate node to linear networks. Because we are interested , the network was made undirected. Next, we define the edge flip score as the minimal number of edges which should be added or removed to convert one network into the other, divided by the total number of edges in both networks. This problem is equivalent to the maximum common edge subgraph problem, a known NP-hard problem with no scalable solutions [@bahiense_maximumcommonedge_2012]. We implemented a branch & bound approach for this problem, by first enumerating all possible edge additions and removals with the minimal number of edges (the edge difference between the two networks) and if none of the new networks was isomorphic, try out all solutions with an additional two edge changes. To further limit the search space, we make sure the degree distributions between the two networks are similar, before assessing whether the two networks are isomorphic using the BLISS algorithm [@junttila_engineeringefficientcanonical_2007], as implemented in the R *igraph* package.

```{r}
add_sfig(
  figure_file("edge_flip_comparison.tmp.pdf", experiment_id="3-score_rules/4-edge_flip"), 
  "edge_flip_comparison", 
  paste0("Edge flip score values for common trajectory topologies, defined in ", ref("sfig", "modulenets")), 
  15, 
  5
)
```

### Aggregation of scores
```{r}
add_sfig(
  figure_file("metric_aggregation_overview_v1.tmp.pdf", experiment_id="evaluation_overview"), 
  "aggregation_of_results", 
  "Aggregation methodology of metric scores.", 
  15, 
  5
)
```

@@

### Pipeline
<!-- Look at dyneval methods to write this paragraph -->

@@

```{r}
qc_references <- readRDS(derived_file("checks.rds", "4-method_characterisation")) %>% 
  .$references %>% 
  str_split("[, \n]+") %>% 
  unlist() %>% 
  unique() %>% 
  discard(is.na) %>%
  glue::collapse(., ";") %>%
  paste0("[", ., "]")
```

@@

### Trajectory types and topology complexity
It's already a bit better, though could be improved still!
Expand, but maybe it should be moved to discussion anyway
ADD CAPTION
😉
TODO
ADD CAPTION
FILL IN SIGNIFICANCE LEVELS
to do: add metric figure
needs to be updated with waypoints
this is not used anymore
TODO
More to add here?
* Look at dynmethods::execute_method to write this paragraph
Are there more exectuion details?
